
# Analytic Hierarchy Process (AHP)

## Prosedur Pengolahan AHP

### Data

```{r}
ahpdata <- read.csv("Data/ahp.csv")
ahpdata
```

### Analisis

#### Faktor


```{r}
# Mendefinisikan faktor
faktor <- c("SAL", "QL", "IW", "LC")
# Menampilkan data frame
faktor_data <- ahpdata[, 2:7]
faktor_data
```

```{r}
# install.packages("ahpsurvey")
library(ahpsurvey)
faktor_data_mat <- ahp.mat(df = faktor_data, faktor, 
                           negconvert = TRUE)
faktor_data_mat
```

```{r}
# Consistency
ri <- ahp.ri(nsims = 10000, dim = 4, seed = 42)
ahp.cr(faktor_data_mat, faktor, ri)
```


```{r}
#Treatement Consistency (Jika Tidak Konsisten)
#faktor_data_mat <- ahp.harker(faktor_data_mat, faktor, iterations = 10, stopcr = 0.1)
#ahp.cr(faktor_data_mat, faktor)
```

The ahp.cr function calculates the consistency ratio of each decision-maker, defined by the following equation:

CR = (λ − n)/((n − 1)(RI))

Where λ is the maximum eigenvalue of the pairwise comparison matrix, n is the number of attributes, and RI is the random index. Following Saaty and Tran (2007), the RI is a function of n and is the consistency ratio of randomly generated pairwise comparison matrices.

Saaty showed that when the CR is higher than 0.1, the choice is deemed to be inconsistent

#### Individual Rangking Faktor

```{r}
library(tidyverse)
library(tibble)
faktor_ind <- ahp.indpref(faktor_data_mat, 
                          faktor, 
                          method = "arithmetic")
round(faktor_ind, 3) %>% rownames_to_column('ID')
```

#### Aggregate Rangking Faktor


```{r}
faktor_agg <- ahp.aggpref(faktor_data_mat, 
                          faktor, 
                          method = "arithmetic", 
                          aggmethod = "arithmetic")
round(faktor_agg, 3) %>% t()
```

```{r}
barplot(faktor_agg,main="Rangking Faktor")
```

```{r}
library(ggplot2)
# Mengubah Cat menjadi factor dengan label yang diinginkan
data = data.frame("Cat"=row.names(data.frame(faktor_agg)),
                  data.frame(faktor_agg))
data$Cat <- factor(data$Cat, 
                   levels = c("SAL", "QL", "IW", "LC"),
                   labels = c("Salary", "Quality of Life", 
                              "Interes in Work", "Location"))
# Mengurutkan
data$warna <- ifelse(data$faktor_agg == 
                       max(data$faktor_agg), 
                     "terbesar", "lainnya")
# Buat grafik batang
ggplot(data, aes(x = Cat, 
                 y = faktor_agg, 
                 fill = warna)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("terbesar" = "#4682B4", 
                               "lainnya" = "#A9A9A9")) +
  theme_minimal() +
  theme(legend.position = "none") +  # Sembunyikan legenda
  labs(
    title = "AHP: Rangking Faktor",
    y = "Skor",
    x = "")
```

#### Alternatif

#### Alternatif untuk Faktor Salary

```{r}
library(dplyr)
alternatif <- c("J1", "J2", "J3")

# Menampilkan data frame
alternatif_data1 <- ahpdata[,8:10]
alternatif1 <- ahp.mat(df = alternatif_data1, 
                       atts = alternatif, 
                       negconvert = TRUE)
alternatif1_agg <- ahp.aggpref(alternatif1, 
                               alternatif, 
                               method = "arithmetic", 
                               aggmethod = "arithmetic")
round(alternatif1_agg, 3) %>% t()
```


```{r}
#Consistency
ri <- ahp.ri(nsims = 10000, dim = 3, seed = 42)
ahp.cr(alternatif1, alternatif, ri)
```

#### Alternatif untuk Faktor Quality of Life


```{r}
alternatif_data2 <- ahpdata[,11:13]
alternatif2 <- ahp.mat(df = alternatif_data2, 
                       atts = alternatif, 
                       negconvert = TRUE)
alternatif2_agg <- ahp.aggpref(alternatif2, 
                               alternatif, 
                               method = "arithmetic", 
                               aggmethod = "arithmetic")
round(alternatif2_agg, 3) %>% t()
```

```{r}
#Consistency
ahp.cr(alternatif2, alternatif, ri)
```

#### Alternatif untuk Faktor Interest in Work

```{r}
alternatif_data3 <- ahpdata[,14:16]
alternatif3 <- ahp.mat(df = alternatif_data3, 
                       atts = alternatif, 
                       negconvert = TRUE)
alternatif3_agg <- ahp.aggpref(alternatif3, 
                               alternatif, 
                               method = "arithmetic", 
                               aggmethod = "arithmetic")
round(alternatif3_agg, 3) %>% t()
```

```{r}
#Consistency
ahp.cr(alternatif3, alternatif, ri)
```

#### Alternatif untuk Faktor Location

```{r}
alternatif_data4 <- ahpdata[,17:19]
alternatif4 <- ahp.mat(df = alternatif_data4, 
                       atts = alternatif, 
                       negconvert = TRUE)
alternatif4_agg <- ahp.aggpref(alternatif4, 
                               alternatif, 
                               method = "arithmetic", 
                               aggmethod = "arithmetic")
round(alternatif4_agg, 3) %>% t()
```

```{r}
#Consistency
ahp.cr(alternatif4, alternatif, ri)
```

#### Gabungan Alternatif

```{r}
alternatif_agg <- cbind(alternatif1_agg,alternatif2_agg,
                        alternatif3_agg,alternatif4_agg) %*% faktor_agg
alternatif_agg
```

```{r}
barplot(t(alternatif_agg) ,main="Rangking Alternatif")
```

```{r}
data = data.frame("Cat"=row.names(data.frame(alternatif_agg)),
                  data.frame(alternatif_agg))
data$Cat <- factor(data$Cat, 
                   levels = c( "J1" , "J2" ,"J3"),
                   labels = c("Job1", "Job2","Job3"))
# Buat grafik batang
data$warna <- ifelse(data$alternatif_agg == max(data$alternatif_agg), 
                     "terbesar", "lainnya")
# Buat grafik batang
ggplot(data, aes(x = Cat, y = alternatif_agg, fill = warna)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("terbesar" = "#4682B4", 
                               "lainnya" = "#A9A9A9")) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(
    title = "AHP: Rangking Alternatif",
    y = "Skor",
    x = "")
```
